---
output: 
  bookdown::html_document2:
    theme: flatly
    css: "style.css"
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: no
    number_sections: true
    code_download: true
    code_folding: show
title: "Day 01"
date: "Start: 22/03/2023"
author:
  name: Joao Malato
  affiliation: Instituto de Medicina Molecular & Immune-Stats
---

# Libraries {-}


```{r}
library(here)
library(data.table)
library(magrittr)
library(ggplot2)
library(scales)
library(colorspace)
library(janitor)
theme_set(theme_bw())
```


# Introduction

- [link](https://www.paulamoraga.com/book-gds/21-spatial-intro.html)

# Types of data

## Areal data

```{r}
library(sf)
library(mapview)
d <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
mapview(d, zcol = "SID74")
```


```{r}
library(spData)
library(ggplot2)
d <- st_read(system.file("shapes/columbus.shp", package = "spData"), quiet = TRUE)
ggplot(d) + geom_sf(aes(fill = INC))
```


```{r}
library(terra)
d <- rast(system.file("ex/elev.tif", package = "terra"))
plot(d)
```

Best packages to use with raster data:
  + `sf` 
  + `terra`


## Geostatistical data

+ Specific (fixed) monitoring points at fixed locations

+ data from the `sp` package

```{r}
library(sp)
library(sf)
library(mapview)

data(meuse)
meuse <- st_as_sf(meuse, coords = c("x", "y"), crs = 28992)
mapview(meuse, zcol = "lead",  map.types = "CartoDB.Voyager")
```

```{r}
# mapview(properties, zcol = "prpsqm")
```


We are essentially measuring price per square meter (Euros/square meter) of a specific set of apartments in Athens, Greece, in 2017.


```{r}
library(malariaAtlas)
d <- getPR(country = "Zimbabwe", species = "BOTH")
autoplot(d)
```


## Point patterns

+ Data points that arise at random
Examples:
  - fires
  - diseases


+ data from `spatstat.data` package

```{r}
library(spatstat)
# plot(spatstat.data::bdspots)
plot(clmfires, use.marks = FALSE, pch = ".")
```

+ We could estimate the intensity of fires for example

```{r}
plot(humberside)
```

+ Look into clusters and infer on the results

# Spatio-temporal data


```{r}
devtools::install_github("Paula-Moraga/SpatialEpiApp")
library(SpatialEpiApp)
library(sf)
library(ggplot2)
library(viridis)

# map
pathshp <- system.file("SpatialEpiApp/data/Ohio/fe_2007_39_county/fe_2007_39_county.shp",
package = "SpatialEpiApp")
map <- st_read(pathshp, quiet = TRUE)

# data
namecsv <- "SpatialEpiApp/data/Ohio/dataohiocomplete.csv"
d <- read.csv(system.file(namecsv, package = "SpatialEpiApp"))

# data are disaggregated by gender and race
# aggregate to get population in each county and year
d <- aggregate(x = d$n, by = list(county = d$NAME, year = d$year), FUN = sum)
names(d) <- c("county", "year", "population")

# join map and data
mapst <- dplyr::left_join(map, d, by = c("NAME" = "county"))

# map population by year
# facet_wrap() splits the data into subsets and create multiple plots
ggplot(mapst, aes(fill = log(population))) + geom_sf() + 
  facet_wrap(~ year, ncol = 7) + scale_fill_viridis("log(population)") +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank())
```



# Spatial functional data

```{r}
library(sf)
library(mapview)
library(geoFourierFDA)
library(rnaturalearth)

# Map Canada
map <- rnaturalearth::ne_states("Canada", returnclass = "sf")

# Coordinates of stations
d <- data.frame(canada$m_coord)
d$location <- attr(canada$m_coord, "dimnames")[[1]]
d <- st_as_sf(d, coords = c("W.longitude", "N.latitude"))
st_crs(d) <- 4326

# Plot Canada map and location of stations
ggplot(map) + geom_sf() + geom_sf(data = d, size = 6) +
  geom_sf_label(data = d, aes(label = location), nudge_y = 2)
```


```{r}
# Temperature of each station over time
d <- data.frame(canada$m_data)
d$time <- 1:nrow(d)

# Pivot data d from wide to long
df <- tidyr::pivot_longer(data = d,
                   cols = names(d)[-which(names(d) == "time")], # columns to pivot into longer format
                   names_to = "variable", # name of the new column with the column names of the original data
                   values_to = "value") # name of the new column with the values of the original data

# Plot temperature of each station over time
ggplot(df, aes(x = time, y = value)) + geom_line(aes(color = variable))
```

---

# Spatial data in R

- [link](https://www.paulamoraga.com/book-gds/22-spatial-spatialdatainR.html)

## Vector data

+ `sf` package is the one we use here

### Shapefile

Vector spatial data such as points, lines and polygons are often represented using a data storage format called shapefile. A shapefile is not a single file but a collection of related files. A shapefile has three mandatory files:

  - `shp`: contains the geometry data,
  - `shx`: positional index of the geometry data that allows to seek forwards and backwards the .shp file,
  - `dbf`: stores the attributes for each shape.

Other files that can form a shapefile include the following:

  - `prj`: plain text file describing the projection,
  - `sbn` and .sbx: spatial index of the geometry data,
  - `shp.xml`: geospatial metadata in XML format.

Therefore, when working with a shapefile, it is important to obtain all files that compose the shapefile and not only the .shp file with the geometry data.

```{r}
library(sf)

# we are reading everything
pathshp <- system.file("shape/nc.shp", package = "sf")

map <- st_read(pathshp, quiet = TRUE)

class(map)


head(map)
```

- The _geometry_ column gives the shape

```{r}
plot(map)
```


## Raster data

To work with regions, divided into rectangles/squares

We use
  - `raster` package (old one)
  - [`terra`](https://rspatial.org/terra) is the new package (the one we shold use)
  - [link](https://rspatial.org/pkg/index.html)


### GeoTIFF


```{r}
library(terra)
pathraster <- system.file("ex/elev.tif", package = "terra")
r <- terra::rast(pathraster)
r
plot(r)
```

## Coordinate Reference Systems (CRS)


### CRS with R

- `sf::st_crs()` 
- `terra::crs()`


```{r}
library(sf)
pathshp <- system.file("shape/nc.shp", package = "sf")
map <- st_read(pathshp, quiet = TRUE)

# Get CRS
# st_crs(map)
# Transform CRS
map2 <- st_transform(map, crs = "EPSG:4326")
# Get CRS
# st_crs(map2)
```



---

# Follow up exercise

## Joining data

```{r}
# Map
library(rnaturalearth)
map <- ne_countries(returnclass = "sf")
plot(map)
```

- `wbstats` package has indicators for each country

```{r}
# Download data from each of the countries
library(wbstats)

# determine pattern
indicators <- wb_search(pattern = "pollution")
d <- wb_data(indicator = "EN.ATM.PM25.MC.M3", start_date = 2016, end_date = 2016)
head(d)
```


```{r}
# Join
# library(dplyr)
# map1 <- left_join(map, d, by = c("iso_a3" = "iso3c"))
map1 <- merge(map, d, by.x = "iso_a3", by.y = "iso3c")

# Plot
library(ggplot2)
library(viridis)
library(ggthemes)
ggplot(map1) + 
  geom_sf(aes(fill = EN.ATM.PM25.MC.M3), color = "white") + 
  colorspace::scale_fill_continuous_sequential(palette = "Purple-Blue") +
  # scale_fill_viridis() +
  theme_map() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  ) +
  # guides(fill = guide_legend(title.position = "top")) +
  coord_sf(crs = "+proj=robin", expand = FALSE)

```


---

# Areal data (follow up)

## Spatial risk modelling

- [link](https://www.paulamoraga.com/book-gds/34-arealdata-modeling-example.html#Spatial_modeling_of_disease_risk_using_areal_data)

- [Example](https://www.envhealthatlas.co.uk/eha/Lung/) used in this exercise to see the geographical patterns of female lung cancer in the UK
  - it compares the risk _relatively_ to the overall risk (absolute) of the country
- [Example 2](https://atlas.cancer.org.au/app)


## Spatial modeling of risk of lung cancer in Pennsylvania


### Data and map

```{r}
library(SpatialEpi)
# load the data
data(pennLC)
class(pennLC)
```

```{r}
head(pennLC$geo)

# data of interest
head(pennLC$data)

# proportions
head(pennLC$smoking)

# spetial polygon with counties of Pensilvania
pennLC$spatial.polygon
```

`pennLC$spatial.polygon` is from package `sp`, we need to transform it into a `sf` file (function _st_as_sf()_)

```{r}
library(sf)
map <- st_as_sf(pennLC$spatial.polygon)
countynames <- sapply(slot(pennLC$spatial.polygon, "polygons"), function(x){slot(x, "ID")})
```


```{r}
map$county <- countynames
head(map)
```

```{r}
library(ggplot2)
ggplot(map) + geom_sf()
```


### Observed cases

```{r}
library(dplyr)
# add the observed number of cases
d <- group_by(pennLC$data, county) %>% summarize(Y = sum(cases))
head(d)
```


### Expected cases

We can now calculate the expected number of cases

```{r}
pennLC$data <- pennLC$data[order(pennLC$data$county, pennLC$data$race, pennLC$data$gender, pennLC$data$age), ]
```


+ The **number of strata** is set to 16 since there are 2 races, 2 genders and 4 age groups for each county ($2 \times 2 \times 4 = 16$).

```{r}
# from package SpatialEpi
E <- expected(population = pennLC$data$population, 
              cases = pennLC$data$cases, 
              n.strata = 16)
E
```

```{r}
d$E <- E

head(d)
```


### Effect of smoking (Smokers proportion)

Adding the smoking data (proportions) into the dataset 

```{r}
d <- left_join(d, pennLC$smoking, by = "county")
```


### Standardized Mortality Ratios

Estimate the Standardized Mortality Ratios

```{r}
d$SMR <- d$Y/d$E

head(d)

d %>% 
  mutate(SMR1 = SMR-1) %>% 
  ggplot(aes(SMR1, reorder(county, SMR))) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_col(aes(fill = SMR1 > 0)) +
  scale_x_continuous(limits = c(-1, 1), expand = expansion(0,0)) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

### Mapping SMR


```{r}
library(dplyr)
map <- left_join(map, d, by = "county")
head(map)
```

```{r}
library(mapview)
library(RColorBrewer)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
pal <- colorspace::sequential_hcl(palette = "Blu-Yl", n = 50)

mapview(
  map,
  zcol = "SMR",
  color = "white",
  alpha.regions = 0.8,
  layer.name = "SMR",
  col.regions = pal,
  map.types = "CartoDB.Positron"
)
```

- We can personalise the map to show certain options only

```{r}
library(mapview)
library(RColorBrewer)
library(leafpop)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
mapviewOptions(fgb = FALSE)

popuptable <-
  leafpop::popupTable(
    dplyr::mutate_if(map, is.numeric, round, digits = 2),
    zcol = c("county", "Y", "E", "smoking", "SMR"),
    row.numbers = FALSE,
    feature.id = FALSE
  )
mapview(
  map,
  zcol = "SMR",
  color = "gray",
  col.regions = pal,
  highlight = leaflet::highlightOptions(weight = 4),
  popup = popuptable
)


```



### Modelling

- Smooth analysis
- 
- uncertainty associated to the covs of the risk

Let $Y_i$ and $E_i$ be the observed and expected number of disease cases, respectively, and let $\theta_i$ be the relative risk for county $i=1,.,n$. The model is specified as follows:

$$Y_i|\theta_i \sim Poisson(E_i\times \theta_i), i=1,.,n,$$

$$\log(\theta_i) = \beta_0 + \beta_1 \times smoking_i + u_i + v_i$$

where $\beta_0$ is the intercept and $\beta_1$ is the coefficient of the covariate smokers proportion. $u_i$ is a structured spatial effect, $u_i | \mathbf{u_i} \sim N(u_{\delta}_{i}, \frac{1}{})$ (intrinsic conditionally autoregressive model (CAR)), and $v_i$ is an unstructured spatial effect, $v_i ??? N(0,1/??v)$.


### Neighbourhood matrix


```{r}
library(spdep)
library(INLA)

nb <- poly2nb(map)
head(nb)
```

```{r}
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
```


### Inference using `INLA`

```{r}
map$re_u <- 1:nrow(map)
map$re_v <- 1:nrow(map)
```

```{r}
formula <- Y ~ smoking + f(re_u, model = "besag", graph = g, scale.model = TRUE) + f(re_v, model = "iid")
# besag is the conditional autorregressive model
```

Then, we fit the model using the _inla()_ function with the default priors in `INLA`. We specify the _formula_, _family_, _data_, and the expected counts ( _E_ ), and set _control.predictor = list(compute = TRUE)_ and _control.compute = list(return.marginals.predictor = TRUE)_ to compute the posterior means of the predictors.

```{r}
# results
res <- inla(
  formula,
  family = "poisson",
  data = map,
  E = E,
  control.predictor = list(compute = TRUE),
  control.compute = list(return.marginals.predictor = TRUE)
)
```


### Results



```{r}
summary(res)
```

```{r}
# look at the estimated risks
res$summary.fitted.values

map$RR <- res$summary.fitted.values[, "mean"] # relative risk
map$LL <- res$summary.fitted.values[, "0.025quant"] # lower limit 95% CI
map$UL <- res$summary.fitted.values[, "0.975quant"] # upper limit 95% CI

# plot(res)
```

### Mapping disease risk

```{r}
library(mapview)
library(RColorBrewer)
library(leafpop)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
mapviewOptions(fgb = FALSE)
mapview(
  map,
  zcol = "RR",
  color = "gray",
  col.regions = pal,
  highlight = leaflet::highlightOptions(weight = 4),
  popup = leafpop::popupTable(
    dplyr::mutate_if(map, is.numeric, round, digits = 2),
    zcol = c("county", "Y", "E", "smoking", "SMR", "RR", "LL", "UL"),
    row.numbers = FALSE,
    feature.id = FALSE
  )
)
```

### Comparing SMR and RR maps


```{r}
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
at <- seq(min(map$SMR), max(map$SMR), length.out = 8)

m1 <- mapview(map, zcol = "SMR", color = "gray", col.regions = pal, at = at)
m2 <- mapview(map, zcol = "RR", color = "gray", col.regions = pal, at = at)
m <- leafsync::sync(m1, m2)
m
```

All areal values have been smoothed into being closer to 1.

### Exceedance probabilities

Question: What is the probability that the risk exceeds X?

```{r}
c <- 1.2
marg <- res$marginals.fitted.values[[51]]
# 1 - probability that the risk is less that c
1 - inla.pmarginal(q = c, marginal = marg)
```


```{r}
library(ggplot2)
marginal <- inla.smarginal(res$marginals.fitted.values[[51]])
marginal <- data.frame(marginal)
ggplot(marginal, aes(x = x, y = y)) + 
  geom_line() +
  labs(x = expression(theta[1]), y = "Density") +
  geom_vline(xintercept = 1.2, col = "black") + 
  theme_bw()
```

```{r}
c <- 1.2
map$exc <- sapply(res$marginals.fitted.values,
FUN = function(marg){1 - inla.pmarginal(q = c, marginal = marg)})
```

```{r}
c <- 1.2
map$exc <- sapply(res$marginals.fitted.values,
FUN = function(marg){1 - inla.pmarginal(q = c, marginal = marg)})
```

```{r}
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
mapview(map, zcol = "exc", color = "gray", col.regions = pal, map.types = "CartoDB.Positron")
```

Answer: Not very likely that these areas will exceed 1.2.



