---
output: 
  bookdown::html_document2:
    theme: flatly
    css: "style.css"
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: no
    number_sections: true
    code_download: true
    code_folding: show
title: "Making maps with R"
date: "Start: 22/03/2023"
author:
  name: Joao Malato
  affiliation: Instituto de Medicina Molecular & Immune-Stats
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE, 
                      fig.align = "center",
                      fig.asp = 0.618,
                      fig.width = 10,
                      dpi = 120, 
                      out.width = "75%")
```

# Description {-}



# Libraries {-}

```{r}
library(here)
library(data.table)
library(magrittr)
library(ggplot2)
library(scales)
library(colorspace)
library(janitor)
theme_set(theme_bw())

```

# Making maps with R

- [link](https://www.paulamoraga.com/book-gds/25-spatial-makingmaps.html)

Data from the North Carolina data

```{r}
library(sf)
nameshp <- system.file("shape/nc.shp", package = "sf")
d <- st_read(nameshp, quiet = TRUE)
# transform
d <- st_as_sf(d)

# infant deaths 1974
d$vble <- d$SID74
# infant deaths 1979
d$vble2 <- d$SID79
head(d)
```

# ggplot2

```{r}
library(ggplot2)
library(viridis)
# ggplot(d) +
#   geom_sf(aes(fill = vble)) +
#   scale_fill_viridis() + 
#   theme_bw()
ggplot(d) +
  geom_sf(aes(fill = vble)) +
  scale_fill_continuous_sequential() + 
  theme_bw() +
  theme(
    legend.position = "top"
  )
```

## plotly

```{r}
library(plotly)
g <- ggplot(d) +
  geom_sf(aes(fill = vble), color = "gray5", size = 0.1) +
  scale_fill_continuous_sequential() + 
  theme_bw()
ggplotly(g)
```


# leaflet

The sf object that we pass to _leaflet()_ needs to have a geographic coordinate reference system (CRS) (EPSG code 4326) indicating latitude and longitude. Here, we use the _st_transform()_ function of `sf` to transform the data d which has CRS given by EPSG code 4267 to CRS with EPSG code 4326.



```{r}
st_crs(d)
```

```{r}
library(leaflet)

pal <- colorNumeric(palette = "YlOrRd", domain = d$vble)

l <- leaflet(d) %>% 
  addTiles() %>%
  addPolygons(
    color = "white",
    fillColor = ~ pal(vble),
    fillOpacity = 0.8
  ) %>%
  addLegend(pal = pal,
            values = ~ vble,
            opacity = 0.8)
l
```


```{r}
l %>% addMiniMap()
```

## Save interactive figure as png

```{r, eval=FALSE}
# Saves map.html

library(htmlwidgets)
saveWidget(widget = l, file = here::here("results/figures", paste(Sys.Date(), "map.html", sep = "_")))

# Takes a screenshot of the map.html created and saves it as map.png
library(webshot)
# webshot::install_phantomjs()
webshot(url = here::here("results/figures", paste(Sys.Date(), "map.html", sep = "_")), file = here::here("results/figures", paste(Sys.Date(), "map.png", sep = "_")))
```


# mapview

```{r}
library(mapview)
mapview(d, zcol = "vble")
```

```{r}
library(RColorBrewer)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
# version 2
mapview(d,
        zcol = "vble",
        map.types = "CartoDB.DarkMatter",
        col.regions = pal)
```

```{r}
# version 3
map1 <- mapview(d, zcol = "vble")
leaflet::addMiniMap(map1@map)
```



# Side by side plots with mapview

```{r}
library(leaflet.extras2) # similar to patchwork
m1 <- mapview(d, zcol = "vble")
m2 <- mapview(d, zcol = "vble2")
m1 | m2
```


# Synchronized maps with leafsync

```{r}
m1 <- mapview(d, zcol = "vble")
m2 <- mapview(d, zcol = "vble2")
m <- leafsync::sync(m1, m2)
m
```


```{r, eval=FALSE}
htmltools::save_html(html = m, 
                     file = here::here("results/figures", paste(Sys.Date(), "map-side-by-side.html", sep = "_")))
```


# `tmap`

```{r}
library(tmap)

# define the mode
tmap_mode("plot") # static
# tmap_mode("view") # interactive

tm_shape(d) + 
  tm_polygons("vble") #+
  # tm_shape(st_centroid(d)) + 
  # tm_dots("vble")
```

# Mobility flows with `flowmapblue`

+ To map mobility data
+ data from cellphone companies

```{r}
# devtools::install_github("FlowmapBlue/flowmapblue.R")
library(flowmapblue)
```

```{r}
# data frame 1
locations <- data.frame(
  id = c(1, 2, 3),
  name = c("New York", "London", "Rio de Janeiro"),
  lat = c(40.713543, 51.507425, -22.906241),
  lon = c(-74.011219, -0.127738, -43.180244)
)
locations

# atadate frame 2
flows <- data.frame(
  origin = c(1, 2, 3, 2, 1, 3),
  dest = c(2, 1, 1, 3, 3 , 2),
  count = c(42, 51, 50, 40, 22, 42)
)
flows
```

```{r}
# final plot of transitions fromand to the 3 destinations
# flowmapblue(locations = locations, flows = flows, mapboxAccessToken = NULL, clustering = TRUE, darkMode = TRUE, animation = TRUE)
```



